---
title: "Dichotomous symptom variables"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
library(effectsize)
library(kableExtra)
```

# Define functions
```{r}
smartT <- function(d,i, dRaw, depVarDF, depVarRawDF, donorsDF) {
  dep <- eval(parse(text=str_c(d)))
  ind <- eval(parse(text=str_c(i)))
  depRaw <- eval(parse(text=str_c(dRaw)))
  categories <- sort(unique(ind))
  cat1N <- sum(!is.na(dep[ind==categories[1]]))
  cat2N <- sum(!is.na(dep[ind==categories[2]]))
  catN <- str_c(cat1N, "/", cat2N)
  categoryNames <- str_c(categories[1],"/",categories[2])
  
  if((cat1N>1) & (cat2N>1)) {
    varEqual <- var.test(dep~ind)$p.value>=0.05
    pval <- t.test(dep~ind, var.equal=varEqual)$p.value
    cohensDstat <- cohens_d(dep~ind, pooled_sd=varEqual)
    cohensD <- cohensDstat$Cohens_d
    cohensDlowCI <- cohensDstat$CI_low
    cohensDhighCI <- cohensDstat$CI_high
    effectSize <- interpret_cohens_d(cohensD)
  } else {
    varEqual<-NA
    pval<-NA
    cohensD<-NA
    cohensDlowCI <- NA
    cohensDhighCI <- NA
    effectSize<-NA
  }
  
  mean1 <- mean(depRaw[ind==categories[1]], na.rm=T)
  sd1 <- sd(depRaw[ind==categories[1]], na.rm=T)
  mean2 <- mean(depRaw[ind==categories[2]], na.rm=T)
  sd2 <- sd(depRaw[ind==categories[2]], na.rm=T)
  
  return(list("categories"=categoryNames,"catN"=catN, "varEqual"=varEqual, "pval"=pval, "cohensD"=cohensD, "cohensDlowCI"=cohensDlowCI, "cohensDhighCI"=cohensDhighCI, "effectSize"=effectSize, "mean1"=mean1, "sd1"=sd1, "mean2"=mean2, "sd2"=sd2))
}
```

The **smartT** (smart t-test) function does the following:

* determines the two categories in the independent variable
* counts the number of observations in each category
* checks that their are sufficient observations for statistical comparison (n>1 in both categories)
* checks to see if the variances of two categories are equal
* if variances are equal, it performs a pooled variance t-test, otherwise it performs a Welch t-test
* computes the effect size (Cohen's D)
* makes a qualitative interpretation of effect size


```{r}
processData <- function(varNames, depVarDF, depVarRawDF, donorsDF) {
  indVar <- unlist(lapply(varNames, rep, length(depVarDF)))
  depVar <- rep(names(depVarDF), length(varNames))
  categories <- NA
  catN <- NA
  varEqual <- NA
  rawP <- NA
  adjP <- NA
  cohensD <- NA
  cohensDlowCI <- NA
  cohensDhighCI <- NA
  effectSize <- NA
  mean1 <- NA
  sd1 <- NA
  mean2 <- NA
  sd2 <- NA
  res <- data.frame(cbind(indVar, depVar, categories, catN, varEqual, rawP, adjP, cohensD, cohensDlowCI, cohensDhighCI, effectSize, mean1, sd1, mean2, sd2))

  indVarT <- str_c("donorsDF$", indVar)
  depVarT <- str_c("depVarDF$", depVar)
  depRawVarT <- str_c("depVarRawDF$", depVar)

  for (i in 1:dim(res)[1]){
    st <- suppressWarnings(smartT(depVarT[i], indVarT[i], depRawVarT[i], depVarDF=depVarDF, depVarRawDF = depVarRawDF, donorsDF = donorsDF))
    res$categories[i] <- st$categories
    res$catN[i] <- st$catN
    res$varEqual[i] <- st$varEqual
    res$rawP[i] <- st$pval
    res$cohensD[i] <- st$cohensD
    res$cohensDlowCI[i] <- st$cohensDlowCI
    res$cohensDhighCI[i] <- st$cohensDhighCI
    res$effectSize[i] <- st$effectSize
    res$mean1[i] <- st$mean1
    res$sd1[i] <- st$sd1
    res$mean2[i] <- st$mean2
    res$sd2[i] <- st$sd2
  }

  res$rawP <- round(as.numeric(res$rawP),4)
  res$adjP <- round(p.adjust(res$rawP, method="BH"),4)
  res$cohensD <- round(as.numeric(res$cohensD),4)
  res$cohensDlowCI <- round(as.numeric(res$cohensDlowCI),4)
  res$cohensDhighCI <- round(as.numeric(res$cohensDhighCI),4)
  res$mean1 <- round(as.numeric(res$mean1),4)
  res$sd1 <- round(as.numeric(res$sd1),4)
  res$mean2 <- round(as.numeric(res$mean2),4)
  res$sd2 <- round(as.numeric(res$sd2),4)

  names(res) <- c("Independent variable", "Dependent variable", "Categories", "Category n", "Equal variances", "Raw p-value", "Adj. p-value (FDR)", "Cohens D", "Cohens D CI low", "Cohens D CI high", "Effect size", "Mean cat 1", "SD cat 1", "Mean cat 2", "SD cat 2")
  
  return(res)
}
```
The **processData** function is a wrapper for the **smartT** function to automate analysis of all variables. It takes four arguments:

* *varNames* - a vector of names of the independent variables
* *depVarDF* - a data.frame of dependent variables (Box-Cox transformed)
* *depVarRawDF* - a data.frame of dependent variables (raw)
* *donorsDF* - a data.frame of donor metadata (including dichotomous symptom variables)

**Interpretation of Cohen's D (Cohen, 1988):**

* d < 0.2 - Very small
* 0.2 <= d < 0.5 - Small
* 0.5 <= d < 0.8 - Medium
* d >= 0.8 - Large

*N.B. Rules apply equally to positive and negative d (i.e., they are given as absolute values).*

# List of dichotomous symptom variables
Dichotomous variables are nominal variables with only two categories.

* Dementia / cognitive impairment
* Depression
* Sleep disturbance
* Aggression
* Hallucinations
* Psychotic symptoms


LB.disease.type has three categories: limbic, neocortical and brainstem. However, we have just one observation for brainstem. If we remove the brainstem observation, we can treat LB.disease.type as a dichotomous variable.

# Load data
***N.B. Raw western blot and puncta data have been transformed using Box-Cox method.***
```{r}
donors <- read.csv("../data/donor_metadata.csv", row.names=1, na.strings="NA")

puncta <- read.csv("../data/Cx43_puncta_per_cell_BoxCox_transformed.csv", row.names=1)
punctaRaw <- read.csv("../data/Cx43 puncta per cell - Sheet1.csv", row.names=1)
punctaRaw <- punctaRaw[c(18:29, 10:17, 1:9, 30:40),]

group <- c(rep("P", 20),rep("C", 20))

wb <- read.csv("../data/WB_protein_levels_BoxCox_transformed.csv", row.names=1)
wb <- wb[c(21:40,1:20),]

wbRaw <- read.csv("../data/WB_protein_levels.csv", row.names=1)
wbRaw <- wbRaw[c(21:40,1:20),]

```

Remove single "brainstem" observation from LB.disease.type, so that we can treat it as a dichotomous variable.
```{r}
donors$LB.disease.type <- ifelse(donors$LB.disease.type=="brainstem", NA, donors$LB.disease.type)
```

# All data
## Donor symptom summary
### Dementia/cognitive impairment
```{r}
tbl <- as.data.frame(table(donors$Dementia...cognitive.impairment))
names(tbl) <- c("Dementia", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Depression
```{r}
tbl <- as.data.frame(table(donors$Depression))
names(tbl) <- c("Depression", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Memory problems
```{r}
tbl <- as.data.frame(table(donors$Memory.problems))
names(tbl) <- c("Memory problems", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:25%;'")
```

### Sleep disturbance
```{r}
tbl <- as.data.frame(table(donors$Sleep.disturbance))
names(tbl) <- c("Sleep disturbance", "Frequency")
knitr::kable(tbl, format="html", table.attr = "style='width:25%;'")
```

### Aggression
```{r}
tbl <- as.data.frame(table(donors$Aggression))
names(tbl) <- c("Aggression", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Hallucinations
```{r}
tbl <- as.data.frame(table(donors$Hallucinations))
names(tbl) <- c("Hallucinations", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Psychotic symptoms
```{r}
tbl <- as.data.frame(table(donors$Psychotic.symptoms))
names(tbl) <- c("Psychotic symptoms", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:25%;'")
```

<!-- ### Gender -->
<!-- ```{r} -->
<!-- tbl <- as.data.frame(table(donors$Gender)) -->
<!-- names(tbl) <- c("Gender", "Frequency") -->
<!-- knitr::kable(tbl, format="html", table.attr="style='width:20%;'") -->
<!-- ``` -->

### Symptom variables for analysis
```{r}
varNames <- c("Dementia...cognitive.impairment", 
              "Depression",
              "Memory.problems",
              "Sleep.disturbance",
              "Aggression",
              "Hallucinations",
              "Psychotic.symptoms")
```

## Protein expression (western blot)
```{r}
res <- processData(varNames = varNames,
                   depVarDF = wb,
                   depVarRawDF = wbRaw,
                   donorsDF = donors)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```


## Cx43_puncta_per_cell
```{r}
res <- processData(varNames = varNames,
                   depVarDF = puncta,
                   depVarRawDF = punctaRaw,
                   donorsDF = donors)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```

# Parkinson's disease

## Donor symptom summary

### LB disease type
```{r}
tbl <- as.data.frame(table(donors$LB.disease.type[group=="P"]))
names(tbl) <- c("LB disease type", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:25%;'")
```

### Dementia/cognitive impairment
```{r}
tbl <- as.data.frame(table(donors$Dementia...cognitive.impairment[group=="P"]))
names(tbl) <- c("Dementia", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Depression
```{r}
tbl <- as.data.frame(table(donors$Depression[group=="P"]))
names(tbl) <- c("Depression", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Memory problems
```{r}
tbl <- as.data.frame(table(donors$Memory.problems[group=="P"]))
names(tbl) <- c("Memory problems", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:25%;'")
```

### Sleep disturbance
```{r}
tbl <- as.data.frame(table(donors$Sleep.disturbance[group=="P"]))
names(tbl) <- c("Sleep disturbance", "Frequency")
knitr::kable(tbl, format="html", table.attr = "style='width:25%;'")
```

### Aggression
```{r}
tbl <- as.data.frame(table(donors$Aggression[group=="P"]))
names(tbl) <- c("Aggression", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Hallucinations
```{r}
tbl <- as.data.frame(table(donors$Hallucinations[group=="P"]))
names(tbl) <- c("Hallucinations", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Psychotic symptoms
```{r}
tbl <- as.data.frame(table(donors$Psychotic.symptoms[group=="P"]))
names(tbl) <- c("Psychotic symptoms", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:25%;'")
```

<!-- ### Gender -->
<!-- ```{r} -->
<!-- tbl <- as.data.frame(table(donors$Gender[group=="P"])) -->
<!-- names(tbl) <- c("Gender", "Frequency") -->
<!-- knitr::kable(tbl, format="html", table.attr="style='width:20%;'") -->
<!-- ``` -->

### Symptom variables for analysis
```{r}
varNamesP <- c("LB.disease.type",
              "Dementia...cognitive.impairment", 
              "Depression",
              "Memory.problems",
              "Sleep.disturbance",
              "Aggression",
              "Hallucinations",
              "Psychotic.symptoms")

donorsP <- donors[group=="P",]
```

## Protein expression (western blot)
```{r}
res <- processData(varNames = varNamesP,
                   depVarDF = wb[group=="P",],
                   depVarRawDF = wbRaw[group=="P",],
                   donorsDF = donorsP)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```

## Cx43_puncta_per_cell
```{r}
res <- processData(varNames = varNamesP,
                   depVarDF = puncta[group=="P",],
                   depVarRawDF = punctaRaw[group=="P",],
                   donorsDF = donorsP)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```


# Control

## Donor symptom summary

### Dementia/cognitive impairment
```{r}
tbl <- as.data.frame(table(donors$Dementia...cognitive.impairment[group=="C"]))
names(tbl) <- c("Dementia", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Depression
```{r}
tbl <- as.data.frame(table(donors$Depression[group=="C"]))
names(tbl) <- c("Depression", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:20%;'")
```

### Memory problems
```{r}
tbl <- as.data.frame(table(donors$Memory.problems[group=="C"]))
names(tbl) <- c("Memory problems", "Frequency")
knitr::kable(tbl, format = "html", table.attr = "style='width:25%;'")
```

### Sleep disturbance
```{r}
tbl <- as.data.frame(table(donors$Sleep.disturbance[group=="C"]))
names(tbl) <- c("Sleep disturbance", "Frequency")
knitr::kable(tbl, format="html", table.attr = "style='width:25%;'")
```

### Aggression
```{r}
tbl <- as.data.frame(table(donors$Aggression[group=="C"]))
names(tbl) <- c("Aggression", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Hallucinations
```{r}
tbl <- as.data.frame(table(donors$Hallucinations[group=="C"]))
names(tbl) <- c("Hallucinations", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:20%;'")
```

### Psychotic symptoms
```{r}
tbl <- as.data.frame(table(donors$Psychotic.symptoms[group=="C"]))
names(tbl) <- c("Psychotic symptoms", "Frequency")
knitr::kable(tbl, format="html", table.attr="style='width:25%;'")
```

<!-- ### Gender -->
<!-- ```{r} -->
<!-- tbl <- as.data.frame(table(donors$Gender[group=="C"])) -->
<!-- names(tbl) <- c("Gender", "Frequency") -->
<!-- knitr::kable(tbl, format="html", table.attr="style='width:20%;'") -->
<!-- ``` -->

### Symptom variables for analysis
```{r}
varNamesC <- c("Depression", "Memory.problems")

donorsC <- donors[group=="C",]
```

## Protein expression (western blot)
```{r}
res <- processData(varNames = varNamesC,
                   depVarDF = wb[group=="C",],
                   depVarRawDF = wbRaw[group=="C",],
                   donorsDF = donorsC)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```

## Cx43_puncta_per_cell
```{r}
res <- processData(varNames = varNamesC,
                   depVarDF = puncta[group=="C",],
                   depVarRawDF = punctaRaw[group=="C",],
                   donorsDF = donorsC)
kable(res, "html") %>% 
  kable_styling() %>%
  scroll_box(width="100%", height="500px")
```


# References
Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd Ed.). New York: Routledge.
```{r}
citation()
citation("stringr")
citation("effectsize")
citation("kableExtra")
```
