---
title: "Donor Metadata: Exploratory Analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
library(ggplot2)
library(viridis)
```

# Define functions
```{r}
swPval <- function(x) {
  shapiro.test(x)$p.value
}
```


# Load data
Read csv file:
```{r}
donors <- read.csv("~/Cam_Dropbox/Nat_Hastings/PD_data/donor_metadata.csv", row.names=1, na.strings="NA")
```
Create variable to represent group:
```{r}
grp <- c(rep("P", 20),rep("C", 20))
grpColor <- ifelse(grp=="C", "blue", "red")
grpPch <- ifelse(grp=="C", 15, 16)
```
Set datatype of columns in donors data.frame:
```{r}
donors$Dementia...cognitive.impairment <- as.factor(donors$Dementia...cognitive.impairment)
donors$Depression <- factor(donors$Depression)
donors$Memory.problems <- as.factor(donors$Memory.problems)
donors$Sleep.disturbance <- as.factor(donors$Sleep.disturbance)
donors$Aggression <- as.factor(donors$Aggression)
donors$Hallucinations <- as.factor(donors$Hallucinations)
donors$Psychotic.symptoms <- as.factor(donors$Psychotic.symptoms)
donors$LB.disease.type <- factor(donors$LB.disease.type)
donors$LB.Braak.stage <- factor(donors$LB.Braak.stage, ordered=TRUE)
donors$SN.depigmentation <- factor(donors$SN.depigmentation, ordered=TRUE)
donors$Amyloid.pathology <- factor(donors$Amyloid.pathology, ordered=TRUE)
donors$Tau.pathology <- factor(donors$Tau.pathology, ordered=TRUE)
donors$Vessel.disease...CAA <- factor(donors$Vessel.disease...CAA, ordered=TRUE)
donors$Age <- as.numeric(donors$Age)
donors$PD.duration <- as.numeric(donors$PD.duration)
donors$Gender <- as.factor(donors$Gender)
```


# Types of variables
The variables in the donor metadata can be broadly divided into two types: categorical (discrete/qualitative) and continuous (quantitative). The categorical variables can be further divided into three subtypes: nominal, dichotomous and ordinal. These are described below.

## Categorical variables

### Nominal
Nominal variables have two or more categories, but do not have an intrinsic order.

* LB disease type

### Dichotomous
Dichotomous variables are nominal variables with only two categories.

* Dementia / cognitive impairment
* Depression
* Sleep disturbance
* Aggression
* Hallucinations
* Psychotic symptoms
* Gender

### Ordinal
Ordinal variables have two or more categories that can be ordered or ranked.

* LB Braak stage
* SN depigmentation
* Amyloid pathology
* Tau pathology
* Vessel disease / CAA

## Continuous variables
* Age
* PD duration

# Differences between PD and control groups

## Categorical variables

### LB disease type
This variable is only applicable to the PD group. There are no data for the control group.
```{r}
df <- data.frame(table(donors$LB.disease.type[grp=="P"]))
names(df) <- c("Disease Type", "Count")
knitr::kable(df)
```

### Dementia / cognitive impairment
```{r}
knitr::kable(table(donors$Dementia...cognitive.impairment, grp))
```

```{r}
df <- data.frame(table(donors$Dementia...cognitive.impairment, grp))
names(df) <- c("Dementia", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Dementia)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Dementia...cognitive.impairment, grp)
```

### Depression
```{r}
knitr::kable(table(donors$Depression, grp))
```

```{r}
df <- data.frame(table(donors$Depression, grp))
names(df) <- c("Depression", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Depression)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Depression, grp)
```


### Sleep disturbance
```{r}
knitr::kable(table(donors$Sleep.disturbance, grp))
```

```{r}
df <- data.frame(table(donors$Sleep.disturbance, grp))
names(df) <- c("Sleep.disturbance", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Sleep.disturbance)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Sleep.disturbance, grp)
```

### Aggression
```{r}
knitr::kable(table(donors$Aggression, grp))
```

```{r}
df <- data.frame(table(donors$Aggression, grp))
names(df) <- c("Aggression", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Aggression)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Aggression, grp)
```

### Hallucinations
```{r}
knitr::kable(table(donors$Hallucinations, grp))
```

```{r}
df <- data.frame(table(donors$Hallucinations, grp))
names(df) <- c("Hallucinations", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Hallucinations)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Hallucinations, grp)
```

### Psychotic symptoms
```{r}
knitr::kable(table(donors$Psychotic.symptoms, grp))
```

```{r}
df <- data.frame(table(donors$Psychotic.symptoms, grp))
names(df) <- c("Psychotic.symptoms", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Psychotic.symptoms)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Psychotic.symptoms, grp)
```

### Gender
```{r}
knitr::kable(table(donors$Gender, grp))
```

```{r}
df <- data.frame(table(donors$Gender, grp))
names(df) <- c("Gender", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Gender)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Gender, grp)
```

### LB Braak stage
```{r}
knitr::kable(table(donors$LB.Braak.stage, grp))
```

```{r}
df <- data.frame(table(donors$LB.Braak.stage, grp))
names(df) <- c("LB.Braak.stage", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=LB.Braak.stage)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$LB.Braak.stage, grp)
```

### SN depigmentation
SN depigmentation only recorded for PD donors. 
```{r}
df <- data.frame(table(donors$SN.depigmentation[grp=="P"]))
names(df) <- c("SN.depigmentation", "Count")
knitr::kable(df)
```

### Amyloid pathology
```{r}
knitr::kable(table(donors$Amyloid.pathology, grp))
```

```{r}
df <- data.frame(table(donors$Amyloid.pathology, grp))
names(df) <- c("Amyloid.pathology", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Amyloid.pathology)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Amyloid.pathology, grp)
```

### Tau pathology
```{r}
knitr::kable(table(donors$Tau.pathology, grp))
```

```{r}
df <- data.frame(table(donors$Tau.pathology, grp))
names(df) <- c("Tau.pathology", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Tau.pathology)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Tau.pathology, grp)
```

### Vessel disease / CAA
```{r}
knitr::kable(table(donors$Vessel.disease...CAA, grp))
```

```{r}
df <- data.frame(table(donors$Vessel.disease...CAA, grp))
names(df) <- c("Vessel.disease...CAA", "Group", "Count")

ggplot(data=df, aes(x=Group, y=Count, fill=Vessel.disease...CAA)) +
  geom_bar(stat="identity") +
  scale_fill_brewer(palette = "Dark2")
```

Fisher's exact test to compare groups:
```{r}
fisher.test(donors$Vessel.disease...CAA, grp)
```


## Continuous variables


### Age
***N.B. age is missing for one donor from the control group***

Boxplot:
```{r}
boxplot(donors$Age ~ grp, ylab="Age", xlab="Group")
```

Quantile-Quantile plots:
```{r}
qqnorm(donors$Age[grp=="C"], main="Control")
qqline(donors$Age[grp=="C"], col="red")
qqnorm(donors$Age[grp=="P"], main="PD")
qqline(donors$Age[grp=="P"], col="red")
```

Shapiro-Wilk normality test applied by group:
```{r}
swPval(donors$Age[grp=="C"])
swPval(donors$Age[grp=="P"])
```

As data for the control group do not follow a normal distribution, we will compare the two groups using the Wilcoxon Rank Sum test.
```{r}
wilcox.test(donors$Age ~grp)
```

### PD duration
PD duration by definition only is applicable to the Parkinson's Disease group.

Boxplot:
```{r}
boxplot(donors$PD.duration)
```

Quantile-Quantile plot:
```{r}
qqnorm(donors$PD.duration)
qqline(donors$PD.duration, col="red")
```

Shapiro-Wilk normality test:
```{r}
swPval(donors$PD.duration)
```


# Any evidence of correlation between variables?
Let's use a scatterplot matrix to look for evidence of a relationship between pairs of ordinal and continuous variables. Key: blue squares = control; red circles = PD. 

Scatterplot matrix of data from all donors.
```{r}
pairs(donors[,c(11:17)], col=grpColor, pch=grpPch)
```

Scatterplot matrix of data from PD group.
```{r}
pairs(donors[grp=="P",c(11:17)], col=grpColor[grp=="P"], pch=grpPch[grp=="P"])
```

Scatterplot matrix of data from control group.
```{r}
pairs(donors[grp=="C",c(13:16)], col=grpColor[grp=="C"], pch=grpPch[grp=="C"])
```

Differences between the two groups are apparent, but no obvious correlation between any pair of variables.