---
title: "Western Blot Data: Analysis of Batch Effects"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
```

# Load data

```{r}
wb <- read.csv("~/Cam_Dropbox/Nat_Hastings/PD_data/WB_protein_levels.csv", row.names=1)

grp <- c(rep("C", 20),rep("P", 20))

plotColours <- gsub("C", "blue", grp)
plotColours <- gsub("P", "red", plotColours)

batch <- c(rep("1", 6), rep("2",6), rep("3",6), rep("4",2))
batch <- rep(batch,2)
```

# Graphical analysis of batch effects
## Boxplots
```{r, results="asis"}
for (i in 1:dim(wb)[2]){
  boxplot(wb[,i]~batch, main=str_c(names(wb)[i]), xlab="Batch", ylab="Protein Expression")
}
```

## Principal Component Analysis (PCA)
Plotting symbol indicates batch. Colour of plotting symbol indicates group (blue=control, red=PD).

### Parietal Cortex
```{r}
allPCA <- prcomp(wb[,c(1:4)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours, pch=batch)
```

### Striatum (Basal Ganglion)
```{r}
allPCA <- prcomp(wb[,c(5:8)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours, pch=batch)
```

### Substantia Nigra (mid brain)
```{r}
allPCA <- prcomp(wb[c(1:3, 5:14,16:40),c(9:12)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours[c(1:3, 5:14,16:40)], pch=batch[c(1:3, 5:14,16:40)])
```



# Statistical analysis of batch effects

## Are data normally distributed within batches?
We use the Shapiro-Wilk test for normality.
```{r}
swPval <- function(x) {
  shapiro.test(x)$p.value
}

sw1 <- apply(wb[batch=="1",], 2, swPval)
sw2 <- apply(wb[batch=="2",], 2, swPval)
sw3 <- apply(wb[batch=="3",], 2, swPval)
sw4 <- apply(wb[batch=="4",], 2, swPval)
```

Adjust p-values to account for multiple tests
```{r}
swRawP <- c(sw1, sw2, sw3, sw4)
swAdjP <- p.adjust(swRawP, method="holm")
sw1adj <- swAdjP[1:12]
sw2adj <- swAdjP[23:24]
sw3adj <- swAdjP[25:36]
sw4adj <- swAdjP[37:48]
```

```{r}
swTable <- data.frame(cbind(sw1adj, sw2adj, sw3adj, sw4adj))
names(swTable) <- c("Batch 1 adj. p", "Batch 2 adj. p", "Batch 3 adj. p", "Batch 4 adj. p")
knitr::kable(swTable)
```

## Compare batches using ANOVA
Since data do not significantly deviate from a normal distribution we will use ANOVA to compare batches.
```{r}
aovP <- function(x, group) {
  summary(aov(x~group))[[1]][[1,'Pr(>F)']]
}
aovBatchRaw <- apply(wb, 2, aovP, group=batch)
aovBatchTable <- data.frame(aovBatchRaw, p.adjust(aovBatchRaw, method="holm"))
names(aovBatchTable) <- c("raw p-value", "adjusted p-value")
knitr::kable(aovBatchTable)
```