---
title: "Western Blot Data: Box-Cox Transformation"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
library(corrplot)
library(MASS)
```


# Define functions
```{r}
checkNormGraph <- function(x, tx, xName){
  cat("\n\n### ", xName,"\n")
  par(mfrow=c(1,2))
  qqnorm(x[1:20], main=str_c("Control raw"))
  qqline(x[1:20], col="red")
  qqnorm(x[21:40], main=str_c("PD raw"))
  qqline(x[21:40], col="red")
  qqnorm(tx[1:20], main=str_c("Control transformed"))
  qqline(tx[1:20], col="red")
  qqnorm(tx[21:40], main=str_c("PD transformed"))
  qqline(tx[21:40], col="red")

  par(mfrow=c(1,2))
  boxplot(x~grp, main="Raw", xlab="Group", ylab="Protein Expression")
  boxplot(tx~grp, main="Transformed", xlab="Group", ylab="Protein Expression")
}

swPval <- function(x) {
  shapiro.test(x)$p.value
}

getBoxCoxLambda <- function(varName, dfName) {
  bc <- eval(parse(text=str_c("boxcox(lm(",varName,"~ 1, data=",dfName,"),plotit=FALSE)")))
  lambda <- bc$x[which.max(bc$y)]
  return(lambda)
}


```

# Load data

```{r}
wb <- read.csv("~/Cam_Dropbox/Nat_Hastings/PD_data/WB_protein_levels.csv", row.names=1)

grp <- c(rep("C", 20),rep("P", 20))
```

# Log transform data

# Box-Cox transformation
We'll start by finding the Box-Cox lambda for each variable. We cannot use this method for **StriBG.GDNF**, because it contains two zeroes.
```{r}
bcLambda <- sapply(names(wb)[c(1:7,9:12)],getBoxCoxLambda,dfName="wb")
```

Perform Box-Cox transformation on each variable (except **StriBG.GDNF**) using lambda calculated in the previous step.
$$
y(\lambda)=
\begin{cases}
\{\frac{y^\lambda-1}{\lambda},\ \text{ if}\ \lambda\neq 0; \\
log\ y,\ \text{if}\ \lambda = 0.
\end{cases}
$$

```{r}
wbT <- wb[,c(1:7,9:12)]
for (i in 1:length(bcLambda)){
  if(bcLambda[i]!=0){
    wbT[,i] <- (wbT[,i]^bcLambda[i]-1)/bcLambda[i]
  }else{
    wbT[,i]<-log(wbT[,i])
  }
}

```

As we can't perform a Box-Cox transformation of **StriBG.GDNF**, we'll use the transformation log(x+1).
```{r}
StriBG.GDNF <- log1p(wb$StriBG.GDNF)
```

Reassemble data.frame:
```{r}
wbT <- cbind(wbT[,1:7], StriBG.GDNF, wbT[,8:11])
```

# Graphical analysis of normality

```{r, results="asis"}
for (i in 1:dim(wb)[2]){
  checkNormGraph(wb[,i], wbT[,i], names(wb)[i])
}

```

# Statistical analysis of normality

We use the Shapiro-Wilk test for normality.
```{r}
swC.raw <- apply(wb[1:20,], 2, swPval)
swPD.raw <- apply(wb[21:40,], 2, swPval)
swC.t <- apply(wbT[1:20,], 2, swPval)
swPD.t <- apply(wbT[21:40,], 2, swPval)
#log2transform <- ((swC.log2>swC.raw)&(swPD.log2>swPD.raw))&(swC.raw<0.05|swPD.raw<0.05)
```

```{r}
#swTable <- data.frame(cbind(swC.raw, swC.log2, swPD.raw, swPD.log2, ifelse(log2transform,"yes","no")))
swTable <- data.frame(cbind(swC.raw, swC.t, swPD.raw, swPD.t, bcLambda))
#names(swTable) <- c("Raw control p", "Transformed control p", "Raw PD p", "Transformed PD p", "Log2 transform?")
names(swTable) <- c("Raw control p", "Transformed control p", "Raw PD p", "Transformed PD p", "Box-Cox Lambda")
knitr::kable(swTable)
```


# Write transformed data to file
```{r}
write.csv(wbT, "~/Cam_Dropbox/Nat_Hastings/PD_data/WB_protein_levels_BoxCox_transformed.csv", row.names=T, quote=F)
```