---
title: "Western Blot Data: Exploratory Analysis"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
library(corrplot)
```


# Define functions
```{r}
checkNormGraph <- function(x, xName){
  cat("\n\n### ", xName,"\n")
  qqnorm(x[1:20], main=str_c("QQ plot of ", xName, " control"))
  qqline(x[1:20], col="red")
  qqnorm(x[21:40], main=str_c("QQ plot of ", xName, " PD"))
  qqline(x[21:40], col="red")
  hist(x[1:20], main=str_c("Histogram of ", xName, " control"), xlab="Protein Expression")
  hist(x[21:40], main=str_c("Histogram of ", xName, " PD"), xlab="Protein Expression")
  boxplot(x~grp, main=str_c("Boxplot of ", xName), xlab="Group", ylab="Protein Expression")
}

swPval <- function(x) {
  shapiro.test(x)$p.value
}
```

# Load data

```{r}
wb <- read.csv("../data/WB_protein_levels.csv", row.names=1)

grp <- c(rep("C", 20),rep("P", 20))
```

# Principal Component Analysis (PCA)

## All data
Data incomplete for rows 4 and 15 (no MidSN), so we will remove these rows
```{r}
plotColours <- gsub("C", "blue", grp)
plotColours <- gsub("P", "red", plotColours)
allPCA <- prcomp(wb[c(1:3, 5:14,16:40),], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours[c(1:3, 5:14,16:40)], pch=grp[c(1:3, 5:14,16:40)])
```

## Parietal Cortex
```{r}
allPCA <- prcomp(wb[,c(1:4)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours, pch=grp)
```

## Striatum (Basal Ganglion)
```{r}
allPCA <- prcomp(wb[,c(5:8)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours, pch=grp)
```

## Substantia Nigra (mid brain)
```{r}
allPCA <- prcomp(wb[c(1:3, 5:14,16:40),c(9:12)], scale=T)
plot(allPCA$x[,c(1,2)], col=plotColours[c(1:3, 5:14,16:40)], pch=grp[c(1:3, 5:14,16:40)])
```


# Do variables follow a normal distribution?

## Graphical analysis of normality

```{r, results="asis"}
for (i in 1:dim(wb)[2]){
  checkNormGraph(wb[,i], names(wb)[i])
}

```

## Statistical analysis of normality

We use the Shapiro-Wilk test for normality.
```{r}
swC <- apply(wb[1:20,], 2, swPval)
swPD <- apply(wb[21:40,], 2, swPval)
```

Adjust p-values to account for multiple tests
```{r}
swRawP <- c(swC, swPD)
swAdjP <- p.adjust(swRawP, method="holm")
swCadj <- swAdjP[1:12]
swPDadj <- swAdjP[13:24]
parametricTest <- swCadj>0.05 & swPDadj>0.05
parametricTestAppropriate <- ifelse(parametricTest, "yes", "no")
```

```{r}
swTable <- data.frame(cbind(swC, swCadj, swPD, swPDadj, parametricTestAppropriate))
names(swTable) <- c("Control raw p", "Control adj. p", "PD raw p", "PD adj. p", "Parametric test appropriate?")
knitr::kable(swTable)
```

# Statistical comparison PD vs control

```{r}
statComp <- function(x, group, paramTest){
  if(paramTest){
    return(t.test(x~group)$p.value)
  }else{
    return(wilcox.test(x~group)$p.value)
  }
}

#rawP <- apply(wb, 2, statComp, group=grp, paramTest=parametricTest)

rawP<-vector()
for (i in 1:dim(wb)[2]){
  rawP <- c(rawP, statComp(x=wb[,i], group=grp, paramTest=parametricTest[i]))
}


adjP <- p.adjust(rawP, method="holm")
testType <- ifelse(parametricTest, "t-test", "Mann-Whitney")
testTable <- data.frame(cbind(testType, rawP, adjP))
names(testTable) <- c("Test", "raw p-value", "adj. p-value")
knitr::kable(testTable)
```

# Correlation between variables
We'll use Spearman's rho statistic which measures the strength and direction of any monotonic association between two variables. *N.B.* relationship does not need to be linear.

```{r}
corMat <- cor(wb, use="pairwise.complete.obs", method="spearman")
corrplot(corMat, method="circle")
knitr::kable(corMat)
```