---
title: "Western Blot Data: Do variables follow a normal distribution?"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time())`"
---

# Load libraries
```{r}
library(stringr)
library(corrplot)
```


# Define functions
```{r}
checkNormGraph <- function(x, log2x, xName){
  cat("\n\n### ", xName,"\n")
  par(mfrow=c(1,2))
  qqnorm(x[1:20], main=str_c("Control raw"))
  qqline(x[1:20], col="red")
  qqnorm(x[21:40], main=str_c("PD raw"))
  qqline(x[21:40], col="red")
  qqnorm(log2x[1:20], main=str_c("Control log2"))
  qqline(log2x[1:20], col="red")
  qqnorm(log2x[21:40], main=str_c("PD log2"))
  qqline(log2x[21:40], col="red")

  par(mfrow=c(1,2))
  boxplot(x~grp, main="Raw", xlab="Group", ylab="Protein Expression")
  boxplot(log2x~grp, main="Log2", xlab="Group", ylab="Protein Expression")
}

swPval <- function(x) {
  shapiro.test(x)$p.value
}
```

# Load data

```{r}
wb <- read.csv("~/Cam_Dropbox/Nat_Hastings/PD_data/WB_protein_levels.csv", row.names=1)

grp <- c(rep("C", 20),rep("P", 20))
```

# Log transform data
We'll use a log2 transformation for all variables except **StriBG.GDNF**. As **StrigBG.GDNF** contains zeroes, we'll use the transformation log2(x+1).
```{r}
log2wb <- log2(wb[,c(1:7,9:12)])
StriBG.GDNF <- log2(wb$StriBG.GDNF+1)
log2wb <- data.frame(cbind(log2wb[,1:7], StriBG.GDNF, log2wb[,8:11]))
rm(StriBG.GDNF)
```

# Graphical analysis of normality

```{r, results="asis"}
for (i in 1:dim(wb)[2]){
  checkNormGraph(wb[,i], log2wb[,i], names(wb)[i])
}

```

# Statistical analysis of normality

We use the Shapiro-Wilk test for normality.
```{r}
swC.raw <- apply(wb[1:20,], 2, swPval)
swPD.raw <- apply(wb[21:40,], 2, swPval)
swC.log2 <- apply(log2wb[1:20,], 2, swPval)
swPD.log2 <- apply(log2wb[21:40,], 2, swPval)
log2transform <- ((swC.log2>swC.raw)&(swPD.log2>swPD.raw))&(swC.raw<0.05|swPD.raw<0.05)
```

```{r}
swTable <- data.frame(cbind(swC.raw, swC.log2, swPD.raw, swPD.log2, ifelse(log2transform,"yes","no")))
names(swTable) <- c("Raw control p", "Log2 control p", "Raw PD p", "Log2 PD p", "Log2 transform?")
knitr::kable(swTable)
```

We'll use a log2 transform if both of the following two conditions are met:
* the original data are not normal for one or both groups (control and PD)
* it improves the distribution of both control and PD data.

it improves the normality of the data for both control and PD groups. 